---
title: "Ecological Networks Lab"
output: html_document
date: '2022-12-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ecological Networks Lab

Session: Afternoon, Day 2 from 3:30 - 6:30 PM

By: Sophia Simon, Daniel Valencia, Kayla Hale, Isidora Ávila-Thieme

Oceans are facing a crisis worldwide, with fish stocks and marine ecosystems collapsing due to anthropogenic stressors including fisheries exploitation and climate change. The field of ecological networks can contribute to understanding how the perturbations caused by those stressors propagate through marine ecosystems by analyzing the structure and dynamics of networks of species interactions.

This aim of this lab is to introduce a code base for analyzing the effects of perturbation from environmental or anthropogenic stressors, that may result in the extinction of marine organisms. Specific focus is placed on how these extinctions may propagate through a food web and impact static network metrics. We extend this analysis to dynamic network metrics in the lab's final reflection, where we also recommend additional resources.

### Part 1 - Introduction to networks approach & basic concepts.

```{r Load packages and visualize data}

#install the R package "NetworkExtinction"

#load packages
library("NetworkExtinction")
library("ggplot2")
library("network")

#load chilean intertidal network
data("chilean_intertidal")

#code to convert between matrix and network formats
intertidal_matrix <- as.matrix(chilean_intertidal)
intertidal_network <- as.network(chilean_intertidal)

```

Below are important definitions relevant to the NetworkExtinction R package workflows (NextR package preprint, Valdovinos 2019 Box 1). Read through these definitions and complete the exercises and reflections that follow each definition.

**Interaction Type** - a link between two nodes reflecting the type of relationship involved. NetworkExtinction handles mutualistic (+/+) and trophic / parasitic (-/+) interaction types. For a more exhaustive overview of interaction types, consult Morales-Castilla et al. (2015).

Reflection 1.1. Interaction type greatly impacts how links between nodes are interpreted biologically and subsequently impact the consequences of loss of connections in extinction cascades. Brainstorm how the loss of links/isolation of nodes relates to extinction in trophic versus mutualistic networks. How might outcomes differ between basal species/consumer nodes and facultative/obligate links?

**Species Richness (S)** -- Total number of species in the network.

```{r Exercise 1.1}

# Calculate the total number of species in the Chilean network
S <- nrow(intertidal_matrix)
```

**Degree** - The number of links that connects a single node to the rest of the networks. Total degree includes in-degrees (e.g. resource links) and out-degrees (e.g. consumer links).

```{r Exercise 1.2}

#Calculate the in-degree, out-degree, and total degree for species 29 in the Chilean intertidal network. 

in_degree <- sum(intertidal_matrix[,29]) #columns, resources of that node
out_degree <- sum(intertidal_matrix[29,]) #rows, consumers of that node
total_degree <- in_degree + out_degree

#repeat for species 64
in_degree <- sum(intertidal_matrix[,64]) #columns, resources of that node
out_degree <- sum(intertidal_matrix[64,]) #rows, consumers of that node
total_degree <- in_degree + out_degree
```

Reflection 1.2. What can in_degree and out_degree tell you about the characteristic of a particular species, or, the positioning of a specific node in the food web?

**Connectance (C)** -- Fraction of potential interactions that are realized, C=L/S^2^ where L is the total number of realized interactions (links connecting species) and is the S^2^ total number of possible links in the network (realized if every species interacts with every other species)

```{r Exercise 1.3}

L <- sum(intertidal_matrix)
C <- L/S^2

```

**Degree Distribution** -- The probability distribution that a node in the network is connected to k other nodes (Estrada, 2007). The `DegreeDistribution()` function calculates the cumulative distribution of the number of links that each species in the food network has (Estrada 2007). Then, the observed distribution is fitted to the exponential, and power law models.

```{r Calculating important metrics}

DegreeDistribution(chilean_intertidal,scale="arithmetic")
```

Reflection 1.3. How do you hypothesize that networks with power law vs exponential degree distributions might respond differently to the removal/extinction of most-connected nodes?

### Part 2. Examples & Interpretation

#### Exercise 2.1 Simulating network extinctions

Most-connected

Random

Specific species... Primary producers, due to low nutrient subsidy. Mollusks, due to ocean acidification

Reflection 2.1. Brainstorm examples of biological mechanisms that would lead to the 3 extinction scenarios considered in this package: extinction of most-connected species, random extinctions, or extinction of specific targeted species.

#### Exercise 2.2 Introducing extinction thresholds

By default, the functions in *NetworkExtinction* assume that, for a secondary extinction to happen, a node needs to lose all connections to its prey. However, one may also want to assume that species are only capable of sustaining existence given a threshold of remaining interaction strengths.

#### Exercise 2.3 Introducing rewiring potential 

Ecological networks aren't static and we should assume that species may shift their connections in response to extinctions of an association/interaction partner. Rewiring processes can be simulated with *NetworkExtinction* using the `RewiringDist` argument.

Below, we call a matrix that contains information about similarities or rewiring potential of species indexed by columns to those indexed by rows.

```{r}

## use the functional group column from species characteristics spreadsheet to inform rewiring? 
```

### Part 3. Independent work

### Part 4. Synthesis & Reflection

#### Reflection 4.1

Reflect on the connection between labs 1 & 2 --- this means directly thinking about the effects of physical oceanographic variables like ocean currents, temperature, salinity, etc. on ecological variables like recruitment, life history, dispersal, species interactions, etc.

One example of an the oceanographic phenomenon that occurs on the coast of Chile and California is upwelling. Upwelling is the wind-driven movement of cold, nutrient-rich water from the ocean depth to its surface, replacing the warm and nutrient-depleted water near the surface. This upward flux of waters results in a persistent supply of nutrients that fuels photosynthesis and accounts for high levels of productivity.

Question 1: Take a moment to hypothesize about the effects of upwelling on ecological variables. Which ecological variables would you expect to be impacted the most, and how might that change propagate through the ecological network?

Question 2: Brainstorm other examples of oceanographic phenomena that may impact ecological variables and how. How do you hypothesize that climate change may alter this oceanographic phenomenon and how might that change propagate through the ecological network?

#### Reflection 4.2
